EDA of Pennsylvania Mathematics Assessments 2011-2012 by Barry Davis
========================================================

```{r echo=FALSE, message=FALSE, warning=FALSE, packages}
# Load all of the packages that you end up using in your analysis in this code
# chunk.

# Notice that the parameter "echo" was set to FALSE for this code chunk. This
# prevents the code from displaying in the knitted HTML output. You should set
# echo=FALSE for all code chunks in your file, unless it makes sense for your
# report to show the code that generated a particular plot.

# The other parameters for "message" and "warning" should also be set to FALSE
# for other code chunks once you have verified that each plot comes out as you
# want it to. This will clean up the flow of your report.

# Environment
library(tidyverse)
library(gridExtra)
library(psych)
library(scales)
library(ggthemes)
library(GGally)
knitr::opts_chunk$set(echo=FALSE)
```

```{r echo=FALSE, Load_the_Data}
#Get the data
math <- read.csv("data/math.csv", stringsAsFactors = FALSE)
```

# Overview

This report explores a dataset containing school district level data on the 
academic achievement results of state assessments in mathematics. We will be 
focusing on Pennsylvania.

[Source](https://catalog.data.gov/dataset/consolidated-state-performance-report-201112/resource/4da8f9de-d265-4e2f-b321-f180052df3f7)

# Univariate Plots Section

```{r echo=FALSE, Univariate_Plots}
# Limit data to PA
math.pa <- math %>% filter(STNAM=="PENNSYLVANIA")

# Str
str(math.pa)
```

We'll need to reshape this data to work with it. Many of the columns represent a
grade level and race (or collection of races) of the students. We will keep only 
the records that represent all students, not divided by race. Then we will reshape 
it for easier comparisons.

```{r warning=FALSE}
# Keep "ALL" as to only look at districts, not race.
math.pa <- math.pa %>% select(leanm11,starts_with("ALL"))

# Rename
names(math.pa) <- c("district","all_num","all_pct_m","3_num_m","3_pct_m",
                    "4_num_m","4_pct_m","5_num_m","5_pct_m","6_num_m","6_pct_m","7_num_m",
                    "7_pct_m","8_num_m","8_pct_m","hs_num_m","hs_pct_m")

# Lowercase for easier comparisons
math.pa$district <- tolower(math.pa$district)


# Split data to stitch together later
math.pa.percent <- math.pa[,c(1,3,5,7,9,11,13,15,17)]
math.pa.num <- math.pa[,c(1,2,4,6,8,10,12,14,16)]

# Rename columns for easy merging
names(math.pa.percent) <- c("district", "all", "third", "fourth", "fifth", "sixth",
                    "seventh", "eighth", "high school")
names(math.pa.num) <- c("district", "all", "third", "fourth", "fifth", "sixth",
                            "seventh", "eighth", "high school")

# Reshape the data
math.pa.percent <- gather(math.pa.percent, "grade", "percent", c(2:9))
math.pa.num <- gather(math.pa.num, "grade", "num_students", c(2:9))

# Merge everything together
pa.math <- merge(math.pa.percent, math.pa.num, by = c("district", "grade"), all.x=TRUE)

# Clean up some un-needed space
rm(math.pa.num, math.pa.percent)

# Get rid of identifiers
pa.math$percent <- gsub("GE", "", pa.math$percent)
pa.math$percent <- gsub("LE", "", pa.math$percent)

# Split up percent into high and low columns incase of ranges.
pa.math$low <- NA
pa.math$high <- NA
pa.math <- pa.math %>% separate(percent, c("low", "high"))
pa.math[pa.math$low=="",]$low <- NA
pa.math[is.na(pa.math$high),]$high <- pa.math[is.na(pa.math$high),]$low
pa.math$low <- as.numeric(pa.math$low)
pa.math$high <- as.numeric(pa.math$high)

# Add an average for the range
pa.math$avg_pct <- (pa.math$low + pa.math$high)/2

# Factors for grade levels
pa.math$grade <- as.factor(pa.math$grade)

# Summary
summary(pa.math)
```

The data was reshaped so we now have 3992 rows and 6 variables. Each observation
represents a grade level from a school district. We have the number of students 
who were assessed, the low and high scores (for schools that report in ranges), 
and the average score of the range.

```{r warning=FALSE}
# Distribution of low scores
ggplot(pa.math, aes(x=low)) +
  geom_histogram(binwidth=5) +
  scale_x_continuous(breaks=seq(0,100,10))
```

We look at the scores including the low value of those with ranges here. We can 
see that the distribution is left skewed and peaks around 80%. It's also 
notable how it steeply drops off just after 90%.

```{r warning=FALSE}

# Distribution of high scores
ggplot(pa.math, aes(x=high)) +
  geom_histogram(binwidth=5) +
  scale_x_continuous(breaks=seq(0,100,10))
```

These are the scores including the high value of those with ranges. I expected 
the distribution to not be dramatically different than the previous plot. I do 
see that the steep dropoff just after 90% is gone here due to using the higher 
value of the ranges. We also peak much closer to 90% than previously.

```{r warning=FALSE}
# Distribution of average scores
ggplot(pa.math, aes(x=avg_pct)) +
  geom_histogram(binwidth=5) +
  scale_x_continuous(breaks=seq(0,100,10))
```

This is probably our best representation of the scores, as we are using the 
individual scores plus the mean of any available ranges. We peak around 85%, but 
see that steep dropoff again just after 90%.

```{r}
tapply(pa.math$avg_pct, pa.math$grade, summary)
```

These summarize the avg_pct variable by grade level. We can see that out high 
school assessments are our worst, and our fourth grade assessments are our best. 
The IQR of high school scores is very low too. It would be interesting to see 
what other data sources could help explain this.

```{r warning=FALSE, message=FALSE}
# Distribution of number of students
ggplot(pa.math, aes(x=num_students)) +
  geom_histogram()
```

We will need to remove some outliers from this.

```{r}
# Better bins
ggplot(pa.math, aes(x=num_students)) +
  geom_histogram(binwidth=50) +
  coord_cartesian(xlim=c(0,15000))
```

While we've zoomed in quite a bit, we still don't have a great representation 
of our data. Let's apply a log10 scale.

```{r warning=FALSE}
# Better bins
ggplot(pa.math, aes(x=num_students)) +
  geom_histogram() +
  scale_x_continuous(trans = "log10", 
                     breaks=trans_breaks("log10", function(x) 10^x), 
                     labels=trans_format("log10", math_format(10^.x)))
```

Here's a better view of our distribution of the number of students in each grade 
level for each district.

# Univariate Analysis

### What is the structure of your dataset?

There are `r unique(pa.math$district) %>% length()` school districts. Each district 
has `r unique(pa.math$grade) %>% length()` grade levels that were tested, giving 
us `r nrow(pa.math)` total rows or observations. The variable grade is a factor 
with the following levels:
* `r levels(pa.math$grade)`

### What is/are the main feature(s) of interest in your dataset?

The main features here are average percent proficiency and grade.

### What other features in the dataset do you think will help support your \
investigation into your feature(s) of interest?

I believe district will be a factor in scores. I think we will see scores that 
are consistent within a district.

### Did you create any new variables from existing variables in the dataset?

I created the grade, low, high, num_students, and avg_pct variables. The grade 
and num_students were created by reshaping the data and condensing the columns 
into row values. Some of these scores were ranges, so I split the range into a 
low and high score, then calculated the range average for avg_pct.

### Of the features you investigated, were there any unusual distributions? \
Did you perform any operations on the data to tidy, adjust, or change the form \
of the data? If so, why did you do this?

The number of students distribution was skewed to the right as a few schools had 
a very large number of students. I created a second plot of that distribution to
 get a better look at the number of students in districts under 5000 students.


# Bivariate Plots Section

```{r echo=FALSE, Bivariate_Plots, message=FALSE, warning=FALSE}
# Scatterplot matrix
ggpairs(subset(pa.math, !is.na(low) & !is.na(high)) %>% select(-district))
```

Here we see the relationship between variables in the dataset. Our low, high, 
and avg_pct are highly correlated, but that's because avg_pct is derived from 
those values. Most of our data is not correlated - but this makes sense. We would 
not expect grade level to have a strong relationship with scores.

```{r warning=FALSE}
# Do we see any pattern from 5000ft view?
ggplot(pa.math, aes(x=avg_pct, y=num_students)) +
  geom_point(alpha=.5)
```

Let's remove the outlier and get a better look.

```{r}
# Same plot, removed outlier
ggplot(subset(pa.math, num_students < 60000), aes(x=avg_pct, y=num_students)) +
  geom_point(alpha=1/20)
```

We need to remove more outliers and zoom in on the bulk of our data.

```{r warning=FALSE}
# Same plot, removed outlier
ggplot(subset(pa.math, num_students < 5000), aes(x=avg_pct, y=num_students)) +
  geom_point(alpha=1/20, position = position_jitter(width = 2)) +
  scale_y_continuous(breaks=seq(0,5000,250))
```

So it doesn't appear there's a correlation with number of students and average 
proficiency. Let's take another closer look, but change our view.

```{r warning=FALSE}
# Size for high schools
ggplot(pa.math %>% filter(grade=="all", num_students<2000), aes(x=num_students, y=avg_pct)) +
  geom_line()
```

We can clearly see here that no pattern emerges when looking at number of students 
and the average proficiency.

```{r warning=FALSE, message=FALSE}
# Size for grades
p <- pa.math %>% group_by(grade) %>% 
  summarise(grade_avg = mean(avg_pct, na.rm=TRUE)) %>% 
  ungroup()
ggplot(p, aes(x=grade, y=grade_avg)) +
  geom_col() +
  scale_x_discrete(limits = c("third", "fourth", "fifth", "sixth", "seventh", 
                              "eighth", "high school", "all"))
```

Here we break down the average proficiency across each grade level for all 
districts. It looks like our third, fourth, and seventh graders are the top 3 
performers, followed closely by sixth grade. Again, we see the large dip in the 
high school assessment scores.

# Bivariate Analysis

### Talk about some of the relationships you observed in this part of the \
investigation. How did the feature(s) of interest vary with other features in \
the dataset?

The set only has limited features, there was no variation.

### Did you observe any interesting relationships between the other features \
(not the main feature(s) of interest)?

There weren't many features outside of the main ones to be used.

### What was the strongest relationship you found?

There was no great relationship, so the strongest was number of students. However, 
this was not a good indicator of proficiency.

# Multivariate Plots Section

```{r}
# Let's see the top 25% districts overall, and look for correlations in number 
# of students and scores
p <- pa.math %>% filter(grade=="all", avg_pct > quantile(pa.math$avg_pct, 
                                                         probs=.75, na.rm=TRUE))
p <- pa.math %>% filter(district %in% p$district)
ggplot(p, aes(x=district, y=avg_pct, size=num_students, color=grade)) +
  geom_point(alpha=.5) +
  theme(axis.text.y = element_text(size=7)) +
  coord_flip()
```

This is the top 25% for overall proficiency average. The number of students, or 
size of the school doesn't seem to correlate well with average proficiency for 
the top schools.

```{r warning=FALSE}
# Let's see this by grade level, filtering down even further to less than 5000
# students
p <- pa.math %>% filter(num_students < 5000, !is.na(avg_pct))
ggplot(p, aes(x=avg_pct, y=num_students)) +
  geom_point(alpha=1/10, position=position_jitter(width=2)) +
  facet_wrap(~grade)
```

Here we take another look at the number of students and average proficiency and 
break it down by grade level. There doesn't appear to be any strong relationship 
here.

```{r}
# Random sample to look for any patterns by district and grade level
set.seed(1979)
sampleDistricts <- sample(pa.math$district, 20, replace=FALSE)
pa.math.samp <- pa.math %>% filter(district %in% sampleDistricts)
ggplot(pa.math.samp, aes(x=grade, y=avg_pct)) +
  geom_col() +
  facet_wrap(~district) +
  scale_x_discrete(limits = c("third", "fourth", "fifth", "sixth", "seventh", 
                              "eighth", "high school", "all")) +
  theme(axis.text.x = element_text(angle=90))
```

This looks at the average proficiency by grade, but split amongst a random 
sample of 20 schools. I wanted to see if and patterns would emerge with random 
selection. 

```{r}
# What about the top 20 overall schools?
top20 <- pa.math %>% filter(grade=="all") %>% 
  arrange(desc(avg_pct)) %>% head(20)
p <- pa.math %>% filter(district %in% top20$district)
ggplot(p, aes(x=grade, y=avg_pct)) +
  geom_col() +
  facet_wrap(~district) +
  scale_x_discrete(limits = c("third", "fourth", "fifth", "sixth", "seventh", 
                              "eighth", "high school", "all")) +
  theme(axis.text.x = element_text(angle=90))
```

Since we did not see any patterns in our random sample, I take a look at the top 
20 schools districts. We do see some consistency here throughout each grade level 
within the districts. We also see a repeating pattern of the high school scores 
being lower in almost all of the above.

```{r}
# Bottom 20 schools
bottom20 <- pa.math %>% filter(grade=="all") %>% arrange(avg_pct) %>% head(20)
p <- pa.math %>% filter(district %in% bottom20$district)
ggplot(p, aes(x=grade, y=avg_pct)) +
  geom_col() +
  facet_wrap(~district) +
  scale_x_discrete(limits = c("third", "fourth", "fifth", "sixth", "seventh", 
                              "eighth", "high school", "all")) +
  theme(axis.text.x = element_text(angle=90))
```

Just as there could be patterns in the top schools, it's possible there are 
patterns in the bottom schools, so this plot looks at the bottom 20 districts. 
We do not see any pattern throughout the grade levels of these poor performing 
school districts.

# Multivariate Analysis

### Talk about some of the relationships you observed in this part of the \
investigation. Were there features that strengthened each other in terms of \
looking at your feature(s) of interest?

The proficiency throughout each grade in a given district varies from school to 
school. There seems to be a dip in the high school scores, which could be due to 
a larger student base compared to each individual grade, the variety of math 
being taught (and understood), and the length of time since using the math skills 
that are assessed. We would need to know more about the assessment.

### Were there any interesting or surprising interactions between features?

No, these were fairly expected. I did think each district would have more consistency 
from grade to grade. The top 20 did have very high scores across the board. 

### OPTIONAL: Did you create any models with your dataset? Discuss the \
strengths and limitations of your model.

I did not create any models as it would not make sense to do so with this data.

------

# Final Plots and Summary

### Plot One
```{r echo=FALSE, Plot_One}
# Let's see the top 25% districts overall, and look for correlations in number 
# of students and scores
p <- pa.math %>% filter(grade=="all", avg_pct > quantile(pa.math$avg_pct, 
                                                         probs=.75, na.rm=TRUE))
p <- pa.math %>% filter(district %in% p$district)
ggplot(p, aes(x=district, y=avg_pct, size=num_students, color=grade)) +
  geom_point(alpha=.5) +
  theme_set(theme_grey()) +
  theme(axis.text.y = element_text(size=7)) +
  coord_flip() +
  scale_color_brewer(palette = "Set1", name="Grade") +
  scale_size(name="Number of Students") +
  labs(title="Top 25% School Districts Average Proficiency", 
       subtitle="By Grade Level and Number of Students",
       x="School DIstrict",y="Average Proficiency")
```

### Description One

There doesn't appear to be many patterns here. The high school scores are some 
of the lowest and we see some grouping of fourth grade around the 95 percentile. 
I used this because if a school has a high overall score, perhaps it can be 
related to the number of students in their district.

### Plot Two
```{r echo=FALSE, Plot_Two, fig.width=9}
# What about the top 20 overall schools?
top20 <- pa.math %>% filter(grade=="all") %>% 
  arrange(desc(avg_pct)) %>% head(20)
p <- pa.math %>% filter(district %in% top20$district)
ggplot(p, aes(x=grade, y=avg_pct)) +
  geom_col() +
  facet_wrap(~district) +
  scale_x_discrete(limits = c("third", "fourth", "fifth", "sixth", "seventh", 
                              "eighth", "high school", "all")) +
  theme_set(theme_foundation()) +
  theme(axis.text.x = element_text(angle=90)) +
  labs(title="Top 20 Schools Average Proficiency by Grade",
       x="Grade Level", y="Average Proficiency")
```

### Description Two

The averages across all grades in these top districts are fairly consistent. 
We do see the drop off in scores in the high school grade levels for most of the 
districts. I wanted to see the top schools and whether they were consistent 
throughout each grade level. It does not appear that one grade level is skewing 
any others. This can be for a number of reasons, like hiring good teachers, 
calibrating teachers on what's being taught, or a large emphasis on taking 
(and studying for) assessment tests.

### Plot Three
```{r echo=FALSE, Plot_Three, fig.width=9}
# Bottom 20 schools
bottom20 <- pa.math %>% filter(grade=="all") %>% arrange(avg_pct) %>% head(20)
p <- pa.math %>% filter(district %in% bottom20$district)
ggplot(p, aes(x=grade, y=avg_pct)) +
  geom_col() +
  facet_wrap(~district) +
  scale_x_discrete(limits = c("third", "fourth", "fifth", "sixth", "seventh", 
                              "eighth", "high school", "all")) +
  theme_set(theme_foundation()) +
  theme(axis.text.x = element_text(angle=90)) +
  labs(title="Bottom 20 Schools Average Proficiency by Grade",
       x="Grade Level", y="Average Proficiency")
```

### Description Three

The bottom schools do not have much consistency across grade levels with a few 
exceptions. Wilkinsburg Borough has a very poor proficiency for high school. If 
the top schools were consistent, I wondered if the bottom schools would show any
consistency. The scores here fluctuate from grade to grade, most of which remains 
below a passing level.

------

# Reflection

The scores within a school district are often less consistent across grade levels 
than I originally anticipated. As someone who grew up in Pennsylvania, it is a bit 
disheartening to see how poor some of our school districts are in mathematics. 

There was a lot of reshaping and cleaning that needed to be done for this dataset. 
I stumbled on the two applications of gather() as well as figuring out how to 
separate a field if I was only interested in one of the returned values.

Future work with this dataset can look at some other information such as taxes. 
It would be interesting to see how each school district performs in relation to 
how much school tax they collect. This would help us see what the residents get 
for their hard earned money. We may be able to look at census data as well for 
median household income to see if that plays a role. Information on Facebook 
would be very interesting if paired with this data. For example, we could look at 
current professions and see what the graduates of these schools are doing.
Would we see more STEM professions from the top districts? What about secondary 
education? How many of the graduates go on to secondary education from the top 
schools versus the bottom schools? 